Dissertation Proposal

### 2023-11-07 proposal.tex
Clean up with helps from Overleaf

for button in button_list:
    button()
\end{lstlisting}
That is, a user would expect to see \verb|0| through \verb|9|. In
fact, however, it prints \verb|9| ten times.

@@

In fact, these problems are not at all limited to students. This
specific looping problem even trips up industrial programmers. The
C\# language \emph{changed} to produce \verb|0| through \verb|9| \cite{lippertClosingLoopVariable2009}. It is now also the focus of a language design
change in Go \cite{chaseFixingLoopsGo2023} for a new kind of looping
construct \cite{coxSpecLessErrorprone2023}.
It continues to trip up programmers
(e.g., \cite{sharveyCreatingFunctionsLambdas2022})
despite being
documented as a ``gotcha'' in Python \cite{reitzHitchhikerGuidePython2016}.
% SK: But not everyone has read the document. I don't think the document is
% a Python official document. Also, this sentence sounds like stating that
% the cited Python programmers are stupid.

@@

SMoL's Lispy syntax is
an artifact of the course (\Cref{s:univ-1-course}) using Racket
\cite{friedmanEssentialsProgrammingLanguages2001, krishnamurthiProgrammingLanguagesApplication2007};
however, it also proved to be pedagogically valuable.
I have found the parentheses useful when discussing
scope in conjunction with local-binding features like \lstinline|let|.
This avoids the various confusing ``variable lifting'' semantics found
in languages like Python (\cite{politzPythonFullMonty2013}; see also
posts like \cite{froadieWhatScopeVariable2022}), where the actual defined range of a
variable is not apparent from the source code.

@@

Students are given a working implementation of SMoL, built
as a \verb|#lang| language inside Racket \cite{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present.
(As in Racket, arguments evaluate left-to-right, to give stateful programs an
unambiguous semantics.) The implementation is available online:
\href{https://github.com/shriram/smol}{https://github.com/shriram/smol}.

@@

@@

@@

@@

@@

In \Cref{s:rel-work}, we discuss several papers that have provided
reports of student misconceptions with different fragments of
SMoL. However, it is difficult to know how comprehensive these
are. While some are unclear on the origin of their programs, they
generally seem to be expert-generated.

@@

Finally, we are inspired by the significant body of education research
on \emph{concept inventories} \cite{hestenesForceConceptInventory1992} (with a growing
number for computer science, as a survey lists \cite{taylorComputerScienceConcept2014}). In
terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question has
one correct answer and several wrong ones. However, the wrong ones are
chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
  misconception the student has}. For instance, if the question is
``What is \verb|sqrt(4)|?'', then \verb|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \verb|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many
  settings. For instance, an educator can use them with clickers to
  get quick feedback from a class. If several students pick a
  specific wrong answer, the educator not only knows they are wrong,
  but also has a strong inkling of \emph{precisely what} misconception
  that group has and can address it directly. I expect our
  instruments to be useful in the same way.}

@@

In Quizius, experts create a prompt; in our case, we asked students to
create small but ``interesting'' programs using the SMoL
language. Quizius shows this prompt to students and gathers their
answers. Each student is then shown a set of programs created by other
students and asked to predict (without running it) the value produced
by the program.\footnote{In the course (\Cref{s:univ-1-course}),
  students were given credit for using Quizius but not penalized for wrong
  answers, reducing their incentive to ``cheat'' by running
  programs. They were also told that doing so would diminish the value of
  their answers. Some students seemed to do so anyway, but most
  honored the directive.} Students are also asked to provide a rationale for why
they think it will produce that output.

@@

After running Quizius in the course (\Cref{s:univ-1-course}),
we took over as experts. Determining which is the right
answer is easy. Where expert knowledge is useful is in
\emph{clustering the rationales}. If all the rationales for a wrong
answer are fairly similar, this is strong evidence that there is a
common misconception that generates it. If, however, there are
multiple rationale clusters, that means the program
is not discriminative enough to distinguish the misconceptions, and it
needs to be further refined to tell them apart. Interestingly, even
the correct answer needs to be analyzed, because sometimes correct
answers do have incorrect rationales (again, suggesting the program
needs refinement to discriminate correct conceptions
from misconceptions).

Prior work using Quizius \cite{saarinenHarnessingWisdomClasses2019} finds that students do author programs
that the experts did not imagine. In our case, we seeded Quizius with
programs from prior papers (\Cref{s:rel-work}), which gives the first
few students programs to respond to. However, we found
that Quizius significantly expanded the scope of our problems and
misconceptions. In our final instrument, most programs were directly
or indirectly inspired by the output of Quizius.

@@

        A reader might think that \verb|sum| takes variable arguments (so
        the program produces \verb|6|), but in fact \verb|...| is a single
        variable, so this produces an arity error.

@@

Due to space limitations, we present the entire instrument in \supplRef{SMoL Quizzes/instrument}.
Here we focus on a few programs where student choices correspond to
misconceptions identified earlier, thereby also showing
that the curated programs are
still effective.
Syntactically, \verb|#t| and \verb|#f| are true and false,
while \verb|#(...)| is a vector. I use a * to indicate the correct
answer (which also matches the implementation's output).

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Questions}                       &
    \textbf{Table of Answers}                    \\\hline
    \lstinputlisting{programs/def_twice.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$59\% & \verb|2 0|             \\
          *32\%       & \verb|Error|           \\
          4\%         & \verb|Other: 2 2|      \\
          3\%         & ``Nothing is printed'' \\
          2\%         & \verb|0 0|             \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/dyn_scope.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \verb|Error| \\
          $^\dag$29\% & \verb|42 #t| \\
          6\%         & Other        \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/reassign.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          *53\%       & \verb|#(6 5) 5|                 \\
          *20\%       & \verb|Error|                    \\
          $^\dag$11\% & \verb|#(6 6) 5|                 \\
          $^\dag$11\% & \verb|#(6 6) 6|                 \\
          5\%         & ``Nothing is printed'' or Other \\
        \end{tabularx}
    }                                            \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizFun. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-fun}
\end{table}
\Cref{t:quizzes-data-fun} lists programs in \quizFun{} that we
consider the most interesting. These data confirm the presence of
scope-related misconceptions:
\begin{enumerate}
  \item At least 59\% of students incorrectly believe that a
        variable can be defined twice in one block.
  \item 29\% of students believe that a variable defined in a function
        will be available in the top-level (or global) environment
        after the function is called. That is, these students may have a
        dynamic scope misconception.
  \item 22\% (11\% + 11\%) of students believe variables themselves
        can be passed as arguments and redefined inside the function.
        They disagree on whether the redefinition persists after the
        function call.
\end{enumerate}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                           &
    \textbf{Table of Answers}                       \\\hline
    \lstinputlisting{programs/circularity.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$50\% & \verb|#(#(2 #(2 3)) #(2 3))| \\
          *29\%       & \verb|x=#(x x)| or
          something similar. Both \verb|(left x)| and \verb|(right x)|
          are \verb|x| itself.                       \\
          $^\dag$16\% & \verb|Error|                 \\
          5\%         & Other                        \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/mvec_in_mvec.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \verb|#(#(100 2 3 4) #(100 2 3 4))| \\
          $^\dag$26\% & \verb|#(#(1 2 3 4) #(100 2 3 4))|   \\
          6\%         & \texttt{Error}                      \\
          4\%         & \verb|#(#(1 2 3 4) #(1 2 3 4))|     \\
          2\%         & Other                               \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/vset_in_let.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & \verb|#(10)|  \\
          $^\dag$36\% & \verb|#(123)| \\
          5\%         & Other         \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/var_as_arg.txt}   & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \texttt{12} \\
          $^\dag$31\% & \texttt{0}  \\
          4\%         & Other       \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/setset.txt}       & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & Other: \texttt{5 7}   \\
          $^\dag$27\% & \texttt{6 7}          \\
          11\%        & \texttt{5 5}          \\
          2\%         & Other: \texttt{Error} \\
          1\%         & Other: \texttt{5 6}   \\
        \end{tabularx}
    }                                               \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizState. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-state}
\end{table}
\Cref{t:quizzes-data-state} lists interesting programs after the
addition of state. These data suggest the students have aliasing-related misconceptions:
\begin{itemize}
  \item Up to 50\% of students think vectors are copied rather than
        aliased.
  \item 16\% of students think trying to construct and print a
        self-referring vector would error. (This program is ambiguous:
        \emph{constructing} works in most SMoL languages, but \emph{printing}
        may well cause a problem. These identified ambiguities are
        addressed in \Cref{s:tutor-refine-program-set}.)
  \item 31\% of students think a variable is aliased by a parameter if
        the two variables have the same name. Perhaps interestingly,
        fewer students (27\%) think the variable aliasing would
        happen if the two variables had different names.
\end{itemize}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                                 &
    \textbf{Table of Answers}                             \\\hline
    \lstinputlisting{programs/fun_returns_lambda.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *82\%       & \verb|3|     \\
          $^\dag$17\% & \verb|Error| \\
          1\%         & Other        \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_1.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *74\%       & \verb|4|            \\
          $^\dag$21\% & \verb|3|            \\
          6\%         & Other: \verb|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_2.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *88\%       & \verb|4|            \\
          $^\dag$10\% & \verb|3|            \\
          3\%         & Other: \verb|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/counter.txt}            & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \verb|1 1 2 3 2| \\
          $^\dag$34\% & \verb|1 1 1 1 1| \\
          4\%         & Other            \\
          1\%         & \verb|1 1 2 3 4| \\
        \end{tabularx}
    }                                                     \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizHof. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-hof}
\end{table}
\Cref{t:quizzes-data-hof} lists interesting programs
after the addition of closures and higher-order functions.
It extends what we saw with loops in
\Cref{s:intro}: that students have misconceptions about their interaction with mutable variables:
\begin{itemize}
  \item 17\% of students think \texttt{lambda} functions can't refer
        to free variables.
  \item 21\% of the students think mutating a variable defined outside
        a lambda can't possibly change the behavior of the lambda.
        Perhaps interestingly, this
        misconception seems to depend on how the lambda is constructed
        (compare the middle two programs).
  \item Student understanding of the
        \lstinline|let|-over-\lstinline|lambda| pattern is weak.
\end{itemize}

@@

        This checks whether $\verb|n| < \verb|3|$, but some presumably
        vocalized it as ``greater than \verb|3|, \verb|n|?''.

@@

\begin{table}[!h]
  \centering\begin{tabularx}{\linewidth}{X l X}
    \textbf{Misconception}                                                     &
    \textbf{Question}                                                          &
    \textbf{Table of Answers}
    \\\hline
    \miscon{CallByRef} Function calls alias variables.                         &
    % mut-vars.not_aliased_by_funarg_2 (35)
    \lstinputlisting{programs/call-by-ref.txt}                                 & {
        \begin{tabularx}{\linewidth}{r X}
          78\%   & \verb|12|    \\
          11\%   & \verb|error| \\
          **10\% & \verb|0|     \\
          1\%    & \verb|23|    \\
        \end{tabularx}
      }
    \\\hline
    \miscon{CallsCopyStructs} Function calls copy data structures.             &
    % vectors2.alias_with_funcall (48)
    \lstinputlisting{programs/calls-copy-structs.txt}
                                                                               & {
        \begin{tabularx}{\linewidth}{r X}
          90\%   & \verb|#(173 0)| \\
          **10\% & \verb|#(1 0)|   \\
        \end{tabularx}
    }                                                                              \\\hline
    \miscon{DeepClosure} Closures copy the \textit{values} of free variables.  &
    % lambda3.lambda_remembers_env (70)
    \lstinputlisting{programs/deep-closure.txt}                                & {
        \begin{tabularx}{\linewidth}{r X}
          86\%   & \verb|4|      \\
          **10\% & \verb|3|      \\
          3\%    & \verb|error|  \\
          1\%    & \verb|lambda| \\
        \end{tabularx}
    }                                                                              \\\hline
    \miscon{DefByRef} Variable definitions alias variables.                    &
    % mut-vars.not_aliased_by_defvar_1 (29)
    \lstinputlisting{programs/def-by-ref.txt}                                  & {
        \begin{tabularx}{\linewidth}{r X}
          85\%   & \verb|0 12|                \\
          **12\% & \verb|0 0|                 \\
          2\%    & \verb|error|               \\
          1\%    & depends on implementation. \\
        \end{tabularx}
      }
    \\\hline
    \miscon{DefOrSet}  Both definitions and variable assignments are
    interpreted as follows: if a variable is not defined in the
    current environment, it is defined. Otherwise, it is mutated to
    the new value.
                                                                               &
    % program: mut-vars.update_undefined (28)
    \lstinputlisting{programs/def-or-set.txt}                                  & {
        \begin{tabularx}{\linewidth}{r X}
          85\%   & \verb|error| \\
          **15\% & \verb|2|     \\
        \end{tabularx}
      }
    \\\hline
    \miscon{DefsCopyStructs} Variable definitions copy structures recursively. &
    % vectors2.alias_with_defvar (47)
    \lstinputlisting{programs/defs-copy-structs.txt}                           & {
        \begin{tabularx}{\linewidth}{r X}
          **67\% & \verb|#(100)| \\
          30\%   & \verb|#(200)| \\
          1\%    & \verb|#(300)| \\
          1\%    & \verb|error|  \\
        \end{tabularx}
    }                                                                              \\\hline
  \end{tabularx}
  \caption{Ground misconceptions identified by the SMoL~Tutor. Answers marked
    with ``**'' represent the misconception. (Part I)}
  \label{t:tutor-grounded-misconceptions-1}
\end{table}

\begin{table}[!h]

@@

          96\%  & \verb|error| \\
          **4\% & \verb|402|   \\

@@

          83\%   & \verb|4|     \\
          **11\% & \verb|error| \\
          4\%    & \verb|2|     \\
          1\%    & \verb|8|     \\

@@

          77\%   & \verb|3|     \\
          **23\% & \verb|error| \\

@@

          76\%   & \verb|3|                   \\
          **14\% & \verb|error|               \\

@@

          1\%    & \verb|+inf|                \\

@@

          67\%   & \verb|#(1 2 3)| \\
          **24\% & \verb|#(1 2 4)| \\

@@

          65\%  & \verb|#0=#(#0# #0#)| (Racket circular object notation) \\
          24\%  & \verb|#(#(2 3) #(2 3))|                                \\
          **6\% & \verb|#(#(2 #(2 3)) #(2 3))|                           \\

@@

\begin{figure}[!h]

@@

How well do our results generalize? Certainly there is reason to
question whether our results would apply to other
populations. \Cref{s:generality-tutor} provides preliminary evidence
that the results are not specific to one institution, and
\Cref{s:rel-work} suggests these issues are widespread. Nevertheless,
much more broad testing is needed to confirm our specific instruments.

>>>>

for button in button_list:
    button()
\end{lstlisting}
That is, a user would expect to see \lstinline|0| through \lstinline|9|. In
fact, however, it prints \lstinline|9| ten times.

@@

In fact, these problems are not at all limited to students. This
specific looping problem even trips up industrial programmers. The
C\# language \emph{changed} to produce \lstinline|0| through \lstinline|9| \cite{lippertClosingLoopVariable2009}. It is now also the focus of a language design
change in Go \cite{chaseFixingLoopsGo2023} for a new kind of looping
construct \cite{coxSpecLessErrorprone2023}.
It continues to trip up programmers
(e.g., \cite{sharveyCreatingFunctionsLambdas2022})
despite being
documented as a ``gotcha'' in Python \cite{reitzHitchhikerGuidePython2016}.
% SK: But not everyone has read the document. I don't think the document is
% a Python official document. Also, this sentence sounds like stating that
% the cited Python programmers are stupid.

@@

SMoL's Lispy syntax is
an artifact of the course (\Cref{s:populations}) using Racket
\cite{friedmanEssentialsProgrammingLanguages2001, krishnamurthiProgrammingLanguagesApplication2007};
however, it also proved to be pedagogically valuable.
I have found the parentheses useful when discussing
scope in conjunction with local-binding features like \lstinline|let|.
This avoids the various confusing ``variable lifting'' semantics found
in languages like Python (\cite{politzPythonFullMonty2013}; see also
posts like \cite{froadieWhatScopeVariable2022}), where the actual defined range of a
variable is not apparent from the source code.

@@

Students are given a working implementation of SMoL, built
as a \lstinline|#lang| language inside Racket \cite{felleisenProgrammableProgrammingLanguage2018}. This language
provides only the defined syntax and semantics of SMoL, with no
(other) Racket features present.
(As in Racket, arguments evaluate left-to-right, to give stateful programs an
unambiguous semantics.) The implementation is available online:
\href{https://github.com/shriram/smol}{https://github.com/shriram/smol}.

@@

\subsection{Multiple Populations}%
\label{s:populations}

@@

\subsection{Find Misconceptions in Student-written Programs}%
\label{s:beyond-experts-blindsopts}

@@

\subsection{Find Misconceptions with Expert-written Programs}

@@

\subsection{Multiple-Syntax}

[TODO]

\subsection{Expected Results}

@@

In \Cref{s:background-misconceptions}, we discuss several papers that have provided
reports of student misconceptions with different fragments of
SMoL. However, it is difficult to know how comprehensive these
are. While some are unclear on the origin of their programs, they
generally seem to be expert-generated.

@@

Finally, we are inspired by the significant body of education research
on \emph{concept inventories} \cite{hestenesForceConceptInventory1992} (with a growing
number for computer science, as a survey lists \cite{taylorComputerScienceConcept2014}). In
terms of mechanics, a concept inventory is just an instrument
consisting of multiple-choice questions (MCQs), where each question has
one correct answer and several wrong ones. However, the wrong ones are
chosen with great care. Each one has been validated so that if a
student picks it, we can quite unambiguously determine \emph{what
  misconception the student has}. For instance, if the question is
``What is \lstinline|sqrt(4)|?'', then \lstinline|37| is probably an
uninteresting wrong answer, but if people appear to confuse
square-roots with squares, then \lstinline|16| would be present as an
answer.\footnote{Concept inventories are thus useful in many
  settings. For instance, an educator can use them with clickers to
  get quick feedback from a class. If several students pick a
  specific wrong answer, the educator not only knows they are wrong,
  but also has a strong inkling of \emph{precisely what} misconception
  that group has and can address it directly. I expect our
  instruments to be useful in the same way.}

@@

In Quizius, experts create a prompt; in our case, we asked students to
create small but ``interesting'' programs using the SMoL
language. Quizius shows this prompt to students and gathers their
answers. Each student is then shown a set of programs created by other
students and asked to predict (without running it) the value produced
by the program.\footnote{In the course (\Cref{s:populations}),
  students were given credit for using Quizius but not penalized for wrong
  answers, reducing their incentive to ``cheat'' by running
  programs. They were also told that doing so would diminish the value of
  their answers. Some students seemed to do so anyway, but most
  honored the directive.} Students are also asked to provide a rationale for why
they think it will produce that output.

@@

After running Quizius in the course (\Cref{s:populations}),
we took over as experts. Determining which is the right
answer is easy. Where expert knowledge is useful is in
\emph{clustering the rationales}. If all the rationales for a wrong
answer are fairly similar, this is strong evidence that there is a
common misconception that generates it. If, however, there are
multiple rationale clusters, that means the program
is not discriminative enough to distinguish the misconceptions, and it
needs to be further refined to tell them apart. Interestingly, even
the correct answer needs to be analyzed, because sometimes correct
answers do have incorrect rationales (again, suggesting the program
needs refinement to discriminate correct conceptions
from misconceptions).

Prior work using Quizius \cite{saarinenHarnessingWisdomClasses2019} finds that students do author programs
that the experts did not imagine. In our case, we seeded Quizius with
programs from prior papers (\Cref{s:background-misconceptions}), which gives the first
few students programs to respond to. However, we found
that Quizius significantly expanded the scope of our problems and
misconceptions. In our final instrument, most programs were directly
or indirectly inspired by the output of Quizius.

@@

        A reader might think that \lstinline|sum| takes variable arguments (so
        the program produces \lstinline|6|), but in fact \lstinline|...| is a single
        variable, so this produces an arity error.

@@

Due to space limitations, we present the entire instrument in \supplRef{SMoL Quizzes/instrument}.
Here we focus on a few programs where student choices correspond to
misconceptions identified earlier, thereby also showing
that the curated programs are
still effective.
Syntactically, \lstinline|#t| and \lstinline|#f| are true and false,
while \lstinline|#(...)| is a vector. I use a * to indicate the correct
answer (which also matches the implementation's output).

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Questions}                       &
    \textbf{Table of Answers}                    \\\hline
    \lstinputlisting{programs/def_twice.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$59\% & \lstinline|2 0|             \\
          *32\%       & \lstinline|Error|           \\
          4\%         & \lstinline|Other: 2 2|      \\
          3\%         & ``Nothing is printed'' \\
          2\%         & \lstinline|0 0|             \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/dyn_scope.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \lstinline|Error| \\
          $^\dag$29\% & \lstinline|42 #t| \\
          6\%         & Other        \\
        \end{tabularx}
    }                                            \\\hline
    \lstinputlisting{programs/reassign.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          *53\%       & \lstinline|#(6 5) 5|                 \\
          *20\%       & \lstinline|Error|                    \\
          $^\dag$11\% & \lstinline|#(6 6) 5|                 \\
          $^\dag$11\% & \lstinline|#(6 6) 6|                 \\
          5\%         & ``Nothing is printed'' or Other \\
        \end{tabularx}
    }                                            \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizFun. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-fun}
\end{table}
\Cref{t:quizzes-data-fun} lists programs in \quizFun{} that we
consider the most interesting. These data confirm the presence of
scope-related misconceptions:
\begin{enumerate}
  \item At least 59\% of students incorrectly believe that a
        variable can be defined twice in one block.
  \item 29\% of students believe that a variable defined in a function
        will be available in the top-level (or global) environment
        after the function is called. That is, these students may have a
        dynamic scope misconception.
  \item 22\% (11\% + 11\%) of students believe variables themselves
        can be passed as arguments and redefined inside the function.
        They disagree on whether the redefinition persists after the
        function call.
\end{enumerate}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                           &
    \textbf{Table of Answers}                       \\\hline
    \lstinputlisting{programs/circularity.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          $^\dag$50\% & \lstinline|#(#(2 #(2 3)) #(2 3))| \\
          *29\%       & \lstinline|x=#(x x)| or
          something similar. Both \lstinline|(left x)| and \lstinline|(right x)|
          are \lstinline|x| itself.                       \\
          $^\dag$16\% & \lstinline|Error|                 \\
          5\%         & Other                        \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/mvec_in_mvec.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \lstinline|#(#(100 2 3 4) #(100 2 3 4))| \\
          $^\dag$26\% & \lstinline|#(#(1 2 3 4) #(100 2 3 4))|   \\
          6\%         & \texttt{Error}                      \\
          4\%         & \lstinline|#(#(1 2 3 4) #(1 2 3 4))|     \\
          2\%         & Other                               \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/vset_in_let.txt}  & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & \lstinline|#(10)|  \\
          $^\dag$36\% & \lstinline|#(123)| \\
          5\%         & Other         \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/var_as_arg.txt}   & {
        \begin{tabularx}{\linewidth}{r X}
          *65\%       & \texttt{12} \\
          $^\dag$31\% & \texttt{0}  \\
          4\%         & Other       \\
        \end{tabularx}
    }                                               \\\hline
    \lstinputlisting{programs/setset.txt}       & {
        \begin{tabularx}{\linewidth}{r X}
          *59\%       & Other: \texttt{5 7}   \\
          $^\dag$27\% & \texttt{6 7}          \\
          11\%        & \texttt{5 5}          \\
          2\%         & Other: \texttt{Error} \\
          1\%         & Other: \texttt{5 6}   \\
        \end{tabularx}
    }                                               \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizState. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-state}
\end{table}
\Cref{t:quizzes-data-state} lists interesting programs after the
addition of state. These data suggest the students have aliasing-related misconceptions:
\begin{itemize}
  \item Up to 50\% of students think vectors are copied rather than
        aliased.
  \item 16\% of students think trying to construct and print a
        self-referring vector would error. (This program is ambiguous:
        \emph{constructing} works in most SMoL languages, but \emph{printing}
        may well cause a problem. These identified ambiguities are
        addressed in \Cref{s:tutor-refine-program-set}.)
  \item 31\% of students think a variable is aliased by a parameter if
        the two variables have the same name. Perhaps interestingly,
        fewer students (27\%) think the variable aliasing would
        happen if the two variables had different names.
\end{itemize}

@@

\begin{table}[t]
  \centering
  \begin{tabularx}{\linewidth}{l X}
    \textbf{Question}                                 &
    \textbf{Table of Answers}                             \\\hline
    \lstinputlisting{programs/fun_returns_lambda.txt} & {
        \begin{tabularx}{\linewidth}{r X}
          *82\%       & \lstinline|3|     \\
          $^\dag$17\% & \lstinline|Error| \\
          1\%         & Other        \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_1.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *74\%       & \lstinline|4|            \\
          $^\dag$21\% & \lstinline|3|            \\
          6\%         & Other: \lstinline|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/fun_and_state_2.txt}    & {
        \begin{tabularx}{\linewidth}{r X}
          *88\%       & \lstinline|4|            \\
          $^\dag$10\% & \lstinline|3|            \\
          3\%         & Other: \lstinline|Error| \\
        \end{tabularx}
    }                                                     \\\hline
    \lstinputlisting{programs/counter.txt}            & {
        \begin{tabularx}{\linewidth}{r X}
          *62\%       & \lstinline|1 1 2 3 2| \\
          $^\dag$34\% & \lstinline|1 1 1 1 1| \\
          4\%         & Other            \\
          1\%         & \lstinline|1 1 2 3 4| \\
        \end{tabularx}
    }                                                     \\\hline
  \end{tabularx}
  \caption{Some questions and student answers from \quizHof. Each
    table row is a program and the (relative) frequencies of
    student answers. Answers that can be considered correct are marked
    with \texttt{*}. Wrong answers that we discuss are
    marked with $^\dag$.}
  \label{t:quizzes-data-hof}
\end{table}
\Cref{t:quizzes-data-hof} lists interesting programs
after the addition of closures and higher-order functions.
It extends what we saw with loops in
\Cref{s:intro}: that students have misconceptions about their interaction with mutable variables:
\begin{itemize}
  \item 17\% of students think \texttt{lambda} functions can't refer
        to free variables.
  \item 21\% of the students think mutating a variable defined outside
        a lambda can't possibly change the behavior of the lambda.
        Perhaps interestingly, this
        misconception seems to depend on how the lambda is constructed
        (compare the middle two programs).
  \item Student understanding of the
        \lstinline|let|-over-\lstinline|lambda| pattern is weak.
\end{itemize}

@@

        This checks whether $\verb|n| < \verb|3|$, but some presumably
        vocalized it as ``greater than \lstinline|3|, \lstinline|n|?''.

@@

\begin{table}[!ht]
  \centering\begin{tabularx}{\linewidth}{X l X}
    \textbf{Misconception}                                                     &
    \textbf{Question}                                                          &
    \textbf{Table of Answers}
    \\\hline
    \miscon{CallByRef} Function calls alias variables.                         &
    % mut-vars.not_aliased_by_funarg_2 (35)
    \lstinputlisting{programs/call-by-ref.txt}                                 & {
        \begin{tabularx}{\linewidth}{r X}
          78\%   & \lstinline|12|    \\
          11\%   & \lstinline|error| \\
          **10\% & \lstinline|0|     \\
          1\%    & \lstinline|23|    \\
        \end{tabularx}
      }
    \\\hline
    \miscon{CallsCopyStructs} Function calls copy data structures.             &
    % vectors2.alias_with_funcall (48)
    \lstinputlisting{programs/calls-copy-structs.txt}
                                                                               & {
        \begin{tabularx}{\linewidth}{r X}
          90\%   & \lstinline|#(173 0)| \\
          **10\% & \lstinline|#(1 0)|   \\
        \end{tabularx}
    }                                                                              \\\hline
    \miscon{DeepClosure} Closures copy the \textit{values} of free variables.  &
    % lambda3.lambda_remembers_env (70)
    \lstinputlisting{programs/deep-closure.txt}                                & {
        \begin{tabularx}{\linewidth}{r X}
          86\%   & \lstinline|4|      \\
          **10\% & \lstinline|3|      \\
          3\%    & \lstinline|error|  \\
          1\%    & \lstinline|lambda| \\
        \end{tabularx}
    }                                                                              \\\hline
    \miscon{DefByRef} Variable definitions alias variables.                    &
    % mut-vars.not_aliased_by_defvar_1 (29)
    \lstinputlisting{programs/def-by-ref.txt}                                  & {
        \begin{tabularx}{\linewidth}{r X}
          85\%   & \lstinline|0 12|                \\
          **12\% & \lstinline|0 0|                 \\
          2\%    & \lstinline|error|               \\
          1\%    & depends on implementation. \\
        \end{tabularx}
      }
    \\\hline
    \miscon{DefOrSet}  Both definitions and variable assignments are
    interpreted as follows: if a variable is not defined in the
    current environment, it is defined. Otherwise, it is mutated to
    the new value.
                                                                               &
    % program: mut-vars.update_undefined (28)
    \lstinputlisting{programs/def-or-set.txt}                                  & {
        \begin{tabularx}{\linewidth}{r X}
          85\%   & \lstinline|error| \\
          **15\% & \lstinline|2|     \\
        \end{tabularx}
      }
    \\\hline
    \miscon{DefsCopyStructs} Variable definitions copy structures recursively. &
    % vectors2.alias_with_defvar (47)
    \lstinputlisting{programs/defs-copy-structs.txt}                           & {
        \begin{tabularx}{\linewidth}{r X}
          **67\% & \lstinline|#(100)| \\
          30\%   & \lstinline|#(200)| \\
          1\%    & \lstinline|#(300)| \\
          1\%    & \lstinline|error|  \\
        \end{tabularx}
    }                                                                              \\\hline
  \end{tabularx}
  \caption{Ground misconceptions identified by the SMoL~Tutor. Answers marked
    with ``**'' represent the misconception. (Part I)}
  \label{t:tutor-grounded-misconceptions-1}
\end{table}

\begin{table}[!ht]

@@

          96\%  & \lstinline|error| \\
          **4\% & \lstinline|402|   \\

@@

          83\%   & \lstinline|4|     \\
          **11\% & \lstinline|error| \\
          4\%    & \lstinline|2|     \\
          1\%    & \lstinline|8|     \\

@@

          77\%   & \lstinline|3|     \\
          **23\% & \lstinline|error| \\

@@

          76\%   & \lstinline|3|                   \\
          **14\% & \lstinline|error|               \\

@@

          1\%    & \lstinline|+inf|                \\

@@

          67\%   & \lstinline|#(1 2 3)| \\
          **24\% & \lstinline|#(1 2 4)| \\

@@

          65\%  & \lstinline|#0=#(#0# #0#)| (Racket circular object notation) \\
          24\%  & \lstinline|#(#(2 3) #(2 3))|                                \\
          **6\% & \lstinline|#(#(2 #(2 3)) #(2 3))|                           \\

@@

\begin{figure}[p]

@@

How well do our results generalize? Certainly there is reason to
question whether our results would apply to other
populations. \Cref{s:generality-tutor} provides preliminary evidence
that the results are not specific to one institution, and
\Cref{s:background-misconceptions} suggests these issues are widespread. Nevertheless,
much more broad testing is needed to confirm our specific instruments.
